---
  title: "Thesis"
author: "Owen Fiore"
date: "10/20/2022"
output: pdf_document
---
  
Reading in the Data
```{r}
library(readxl)
times <- read_excel("C:/Users/ofior/Documents/Thesis/Data/ThesisData.xlsx")
View(times)
colnames(times) <- c("Year", "Type", "Time", "RxnTime", "Gender")
```


t test comparing 2019 and 2022
```{r}
t.test(times$RxnTime[times$Year==2019],times$RxnTime[times$Year==2022])
```


Creates a Boxplot comparing 2019 and 2022
```{r}
boxplot(times$RxnTime[times$Year==2019],times$RxnTime[times$Year==2022])
```

Analysis of Variance, looking only at Year
```{r}
model <- aov(RxnTime ~ Year, data = times)
summary(model)
```

Creates a new data set of just finals and looks at the data from that
```{r}
finals <- subset(times, Type == "F")
boxplot(RxnTime ~ Year, data = finals, xlab = "Years")

```



```{r}
library(dplyr)
finals.aov <- aov(RxnTime ~ as.factor(Year), data = finals)
summary(finals.aov)
tuk.final <- TukeyHSD(finals.aov, conf.level = .95)
tkf <- as.data.frame(tuk.final[1])
tkf %>%filter(as.factor.Year..p.adj < 0.05)
#plot(tuk.final)
```


```{r}
residuals <- resid(finals.aov)
plot(finals$RxnTime, residuals, xlab="RxnTime", ylab="Residuals")
abline(0,0)
```

```{r}
library(lme4)
finals.lme <- lmer(RxnTime ~ (1|Year), data = finals)
summary(finals.lme)
qqnorm(resid(finals.lme))
hist(resid(finals.lme))
qqnorm(ranef(finals.lme))
pnorm(0.1, mean = .148883, sd = sqrt(8.648*10^(-5)+3.719*10^(-4)))
```
```{r}
library(lme4)
semi <- subset(times, Type == "S")
finals.lme <- lmer(RxnTime ~ (1|Year), data = semi)
summary(finals.lme)
qqnorm(resid(finals.lme))
hist(resid(finals.lme))
qqnorm(ranef(finals.lme))
```
```{r}
library(lme4)
heat <- subset(times, Type == "H")
finals.lme <- lmer(RxnTime ~ (1|Year), data = heat)
summary(finals.lme)
qqnorm(resid(finals.lme))
hist(resid(finals.lme))
qqnorm(ranef(finals.lme))
str(ranef(finals.lme))
hist(ranef(finals.lme)$Year[,1])
qqnorm(ranef(finals.lme)$Year[,1])
qqline(ranef(finals.lme)$Year[,1])
```
```{r}
pnorm(0.1, mean = .157750, sd = sqrt(.0001228+0.0005937))
qnorm(0.01, mean = .157750, sd = sqrt(.0001228+0.0005937))
```

```{r}
my_fit <- lm(RxnTime ~ 1 , data = finals)
summary(my_fit)
```


```{r}
anova(finals.lme, my_fit)
```

```{r}
old <- subset(times, Year < 2020)
old.lme <- lmer(RxnTime ~ (1|Year), data = old)
summary(old.lme)
old_fit <- lm(RxnTime ~ 1 , data = old)
anova(old.lme, old_fit)


```

```{r}
model_coefs <- coef(old.lme)$Year

model_coefs
```


Plotting LME
```{r}
library(ggplot2)
model_coef_plot <- ggplot(data = old, 
                          mapping = aes(x = Time, 
                                        y = RxnTime, 
                                        colour = Year)
) +
  scale_y_continuous(limits = c(180, 1020)) +
  scale_x_continuous(breaks = seq(1:10) - 1) +
  theme(legend.position = "top")

model_coef_plot
```



Creates a new data set of just semifinals and looks at the data from that
```{r}
semi <- subset(times, Type == "S")

semi.fit <- lm(RxnTime ~ Year, data = semi)
summary(semi.fit)
```

```{r}

boxplot(RxnTime ~ Year, data = semi, xlab = "Years")
```

```{r}

semi.aov <- aov(RxnTime ~ as.factor(Year), data = semi)
summary(semi.aov)
tuk.semi <- TukeyHSD(semi.aov, conf.level = .95)
tks <- as.data.frame(tuk.semi[1])
tks %>%filter(as.factor.Year..p.adj < 0.05)

```
```{r}
residuals.semi <- resid(semi.aov)
plot(semi$RxnTime, residuals.semi, xlab="RxnTime", ylab="Residuals")
abline(0,0)
## Off by 1 error
```

Creates a new data set of just heats and looks at the data from that
```{r}
heat <- subset(times, Type == "H")

heat.fit <- lm(RxnTime ~ Year, data = heat)
summary(heat.fit)
```

```{r}

boxplot(RxnTime ~ Year, data = heat, xlab = "Years")
```

```{r}

heat.aov <- aov(RxnTime ~ as.factor(Year), data = heat)
summary(heat.aov)
tuk.heat <- TukeyHSD(heat.aov, conf.level = .95)
tkh <- as.data.frame(tuk.heat[1])
tkh %>%filter(as.factor.Year..p.adj < 0.05)

```


t test comparing men and women times in 2022
```{r}
men <- subset(times, Gender == "M")
women <- subset(times, Gender == "F")
t.test(men$RxnTime, women$RxnTime)

```

Scatterplot of Data
```{r}
plot(finals$Year, finals$RxnTime)
```
```{r}
ninetynine <- times$RxnTime[times$Year == "1999"]
one <- times$RxnTime[times$Year == "2001"]
three <- times$RxnTime[times$Year == "2003"]
five <- times$RxnTime[times$Year == "2005"]
seven <- times$RxnTime[times$Year == "2007"]
five <- times$RxnTime[times$Year == "2005"]
thirteen <- times$RxnTime[times$Year == "2013"]
two <- times$RxnTime[times$Year == "2022"]
```

```{r}
plot(times$RxnTime, times$Time)
plot(times$RxnTime[times$Time < 15], times$Time[times$Time < 15], xlab = "RxnTime", ylab = "Total Time")
```

